---
title: "RIPPLE benchmarks: null false positive rate, power, and runtime"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RIPPLE benchmarks: null false positive rate, power, and runtime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5, out.width = "100%")
library(ripple)
library(data.table)
library(ggplot2)
benchmarks <- readRDS(system.file("extdata/benchmarks_current_results.rds",
                                package = "ripple", mustWork = TRUE))
pct <- function(x, digits = 1) paste0(formatC(100 * x, format = "f", digits = digits), "%")
summarize_null <- function(x, groups = "n_samples") {
  as.data.table(x)[, .(runs = .N, false_calls = sum(n_sig_fdr),
    tests = sum(n_genes_tested), pooled_fpr = sum(n_sig_fdr) / sum(n_genes_tested),
    empirical_fdr = mean(n_sig_fdr > 0)), by = groups]
}
```

This vignette presents RIPPLE's synthetic benchmarks for null false calls,
gradient recovery, and runtime. The five main sections correspond to Figure
S1A–E: Poisson null, overdispersed null, sign-gate ablation, power at matched
expression, and runtime. Tables and plots use the latest version.
Optional confounder models, query-location permutations, and pathway or ligand-receptor
analyses are not evaluated here.

## Synthetic data design

Statistical benchmarks use a 500 µm field with 30 Tumor cells in a central
disc of radius 60 µm, 120 target T cells distributed across the field, and
50 Fibroblasts in a ring at 100–200 µm from the center. Null datasets have no
planted distance dependence. The power benchmark plants negative slopes in
five T-cell genes, giving higher expression near Tumor cells.

Two error measures are reported separately:

- **Pooled per-gene FPR:** total false discoveries divided by total tested
  null genes across simulations.
- **Empirical FDR under the complete null:** the fraction of simulations with
  any discovery. When all genes are null, every discovery is false, so this
  also equals the empirical family-wise error rate.

Calls use `fisher_fdr < 0.05`. A low pooled per-gene FPR does not by itself
establish 5% FDR control.

## 1. Poisson null (S1A)

Each dataset contains 500 background genes, with 50 simulations at each of
N = 3, 5, and 10 biological replicates: 25,000 null-gene tests per sample
size, or 75,000 overall. The default sign gate requires agreement across all
valid replicates.

```{r poisson-null, fig.cap = "Pooled per-gene FPR after the strict sign gate and BH adjustment. Zero means no false calls were observed in these simulations."}
null <- summarize_null(benchmarks$null_strict)
ggplot(null, aes(factor(n_samples), pooled_fpr)) +
  geom_point(size = 3, colour = "#377EB8") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.001),
                     limits = c(0, NA)) +
  labs(x = "Biological replicates (N)", y = "Pooled per-gene FPR") +
  theme_ripple(base_size = 12)
knitr::kable(null[, .(N = n_samples, Runs = runs, Tests = tests,
  `False calls` = false_calls, `Pooled FPR` = pct(pooled_fpr, 3),
  `Empirical FDR` = pct(empirical_fdr))])
```

The corrected runs produced one false call at N = 3 and none at N = 5 or
10, giving pooled FPRs of 0.004%, 0%, and 0%. These data follow the Poisson
model used for fitting.

### Relaxing the sign gate

The same Poisson-null datasets were also evaluated at `sign_consistency = 0.75`.
The post-BH false-call counts were unchanged in these runs.
The default remains 1.0; relaxation is suggested only for designs with at
least six replicates and should be reported explicitly. N = 5 is included
below as a benchmark comparison. At N = 3, both thresholds require
unanimous signs.

```{r relaxed-null}
relaxed <- summarize_null(benchmarks$null_relaxed)
knitr::kable(relaxed[, .(N = n_samples, `False calls` = false_calls,
  `Pooled FPR` = pct(pooled_fpr, 3), `Empirical FDR` = pct(empirical_fdr))],
  caption = "Poisson-null results with sign consistency 0.75.")
```

## 2. Overdispersed null (S1B)

This benchmark uses 50 background genes and 50 simulations for each sample
size and variance-to-mean ratio (1, 1.5, 2, and 3), giving 600 datasets.
Counts are negative binomial above ratio 1, while the fitted model remains
Poisson. The plot shows pooled per-gene FPR, as in S1B.

```{r overdispersed-null, fig.cap = "Pooled per-gene FPR under increasing count overdispersion, using the unchanged Poisson model and strict sign gate."}
nb <- summarize_null(benchmarks$null_nb, c("dispersion", "n_samples"))
ggplot(nb, aes(dispersion, pooled_fpr, colour = factor(n_samples))) +
  geom_line() + geom_point(size = 2.5) +
  scale_x_continuous(breaks = c(1, 1.5, 2, 3)) +
  scale_y_continuous(labels = scales::label_percent(), limits = c(0, NA)) +
  scale_colour_brewer(palette = "Set1", name = "N") +
  labs(x = "Simulated variance-to-mean ratio", y = "Pooled per-gene FPR") +
  theme_ripple(base_size = 12)
knitr::kable(nb[order(dispersion, n_samples), .(
  `Variance / mean` = dispersion, N = n_samples,
  `Pooled FPR` = pct(pooled_fpr, 2), `Empirical FDR` = pct(empirical_fdr))])
```

At ratio 3, pooled FPRs were 4.8%, 1.2%, and 0.24% for N = 3, 5, and 10.
However, empirical FDR was 80%, 44%, and 10%, respectively. The low per-gene
rates therefore do not demonstrate 5% FDR control under overdispersion.

## 3. Sign-gate ablation (S1C)

This separate simulation exemplifies the combination step. It uses
200,000 sets of five independent uniform p-values and independent random
signs, without simulating cells or fitting GLMs. Under these assumptions,
the probability of unanimous signs is `2 * 0.5^N`, or 1/16 at N = 5.
This factor applies to rejection at a fixed unadjusted Fisher threshold;
it is not a general guarantee for fitted tissue data or post-BH FDR.

```{r sign-ablation, fig.cap = "Idealized Fisher rejection rates with and without the strict sign gate. No BH adjustment is applied in this ablation."}
set.seed(2026)
n_sets <- 200000L
n_reps <- 5L
p <- matrix(runif(n_sets * n_reps), nrow = n_sets)
s <- matrix(sample(c(-1L, 1L), n_sets * n_reps, replace = TRUE), nrow = n_sets)
fisher <- pchisq(-2 * rowSums(log(pmax(p, 1e-15))),
                 df = 2 * n_reps, lower.tail = FALSE)
pass <- abs(rowSums(s)) == n_reps
ablation <- rbindlist(lapply(10^seq(-4, log10(0.5), length.out = 25), function(a) {
  data.table(alpha = a, rate = c(mean(fisher < a), mean(fisher < a & pass)),
             gate = c("Without sign gate", "With sign gate"))
}))
ggplot(ablation, aes(alpha, rate, colour = gate)) +
  geom_line() + geom_point(size = 1.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", colour = "grey50") +
  labs(x = "Unadjusted Fisher threshold", y = "Rejection rate", colour = NULL) +
  theme_ripple(base_size = 12) + theme(legend.position = "bottom")
```

## 4. Power at matched expression (S1D)

This shows expected expression at three counts per planted gene per
target cell in every sample. Tissue layouts are shared across slopes, and
the first three and five samples are subsets of the ten-sample dataset.
For capped distance `d* = min(d, 200)`, the expected count is
`3 * exp(beta * d*) / mean(exp(beta * d*))`, with the mean taken over
that sample's 120 target cells.

Each cell has 500 counts sampled from a multinomial distribution. There are
five planted genes, 45 constant-background genes, and one balancing feature
that carries the remaining counts. This keeps the observed-total offset
equal to the true exposure.
Six slopes (|β| = 0.0005, 0.001, 0.002, 0.005, 0.01, and 0.02 per µm),
three sample sizes, and 30 simulations per combination give 540 datasets.
Both sign gates use identical per-sample fits. Recovery counts all five
planted genes, including any removed by the expression filter. All planted
genes passed the unchanged filter, without priority-gene rescue.

```{r matched-power, fig.width = 5.5, fig.height = 5.5, fig.cap = "Recovery at matched expected expression and sampling. The displayed range matches S1D; the complete simulated range is tabulated below."}
power <- as.data.table(benchmarks$power_matched)[,
  .(recovery = mean(tp / 5)), by = .(beta, n_samples, gate)]
power[, gate_label := factor(gate, levels = c(1, 0.75),
                            labels = c("Strict (1.0)", "Relaxed (0.75)"))]
ggplot(power[abs(beta) >= 0.002 & abs(beta) <= 0.01],
       aes(abs(beta), recovery, colour = factor(n_samples),
           linetype = gate_label, group = interaction(n_samples, gate_label))) +
  geom_line(linewidth = 0.8) + geom_point(size = 2.5) +
  scale_x_log10(breaks = c(0.002, 0.005, 0.01), labels = c("0.002", "0.005", "0.010")) +
  scale_y_continuous(labels = scales::label_percent(), limits = c(0.6, 1),
                     breaks = seq(0.6, 1, 0.1)) +
  scale_colour_brewer(palette = "Set1", name = "Replicates") +
  scale_linetype_manual(values = c("solid", "dashed"), name = "Sign gate") +
  labs(x = expression("Absolute planted gradient score (per " * mu * "m)"),
       y = "Planted gradients recovered") +
  theme_ripple(base_size = 12) +
  theme(legend.position = "bottom", legend.box = "vertical")
power_table <- dcast(power, beta + n_samples ~ gate_label, value.var = "recovery")
knitr::kable(power_table[order(abs(beta), n_samples), .(
  `|beta| per µm` = abs(beta), N = n_samples,
  Strict = pct(`Strict (1.0)`), Relaxed = pct(`Relaxed (0.75)`))])
```

## 5. Runtime and memory (S1E)

Runtime measures `run_ripple()` separately from data generation, with three
timing runs per dataset size. Runs were performed sequentially on a laptop
Windows machine with an Intel Core Ultra 7 165H processor (16 cores,
22 logical processors), using R 4.5.1. `OMP_NUM_THREADS`,
`OPENBLAS_NUM_THREADS`, and `MKL_NUM_THREADS` were each set to one.

Memory is the sum of the maximum-used R heap values reported by `gc()`;
it is not operating-system peak resident memory.

```{r runtime, fig.width = 7, fig.height = 4}
rt <- as.data.table(benchmarks$runtime)[, .(
  seconds = median(t_ripple), heap_mb = median(peak_mb)),
  by = .(size, n_cells, n_genes, n_samples)]
setorder(rt, n_cells)
knitr::kable(rt[, .(Cells = n_cells, Genes = n_genes, N = n_samples,
  `Median time (s)` = round(seconds, 1), `Median peak R heap (MB)` = round(heap_mb))])
rt[, size := factor(size, levels = c("small", "medium", "large"))]
rt_long <- melt(rt, id.vars = "size", measure.vars = c("seconds", "heap_mb"))
rt_long[, metric := factor(variable, levels = c("seconds", "heap_mb"),
                          labels = c("Runtime (s)", "Peak R heap (MB)"))]
ggplot(rt_long, aes(size, value)) +
  geom_col(fill = "#377EB8", width = 0.65) +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = "Dataset size", y = NULL) + theme_ripple(base_size = 12)
```

## Reproducing the results

The vignette renders from the bundled `benchmarks_current_results.rds`, so
building it does not rerun the tissue simulations. The sign-gate ablation
is generated directly in its seeded code chunk.

The scripts below reproduce the simulation
designs. You can run each in a fresh R process from the repository root. They write
to `data-raw/benchmarks/results/`. Runtime measurements depend on the machine and
thread settings.

```{bash rerun, eval = FALSE}
Rscript data-raw/benchmarks/bench_null_hires.R
Rscript data-raw/benchmarks/bench_null_hires_relaxed.R
Rscript data-raw/benchmarks/bench_null_nb.R
Rscript data-raw/benchmarks/bench_power_matched.R 3
Rscript data-raw/benchmarks/bench_power_matched.R 5
Rscript data-raw/benchmarks/bench_power_matched.R 10
Rscript data-raw/benchmarks/bench_runtime.R
```

Run `export_vignette_cache.R` with the benchmark results directory and the
`power_matched` directory to update the bundled tables.

## Session info

This is the vignette-rendering session. Benchmark runs used the R version
and settings described above.

```{r session-info}
sessionInfo()
```
