---
title: "RIPPLE applied to CosMx NSCLC: a worked example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RIPPLE applied to CosMx NSCLC: a worked example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%"
)
```

```{r libraries, message = FALSE}
library(ripple)
library(data.table)
library(ggplot2)
library(ggrepel)
library(patchwork)
```

## Overview

This vignette walks through RIPPLE on the **CosMx SMI NSCLC dataset**
from [He et al. (2022)](https://doi.org/10.1038/s41587-022-01483-z).
765,771 cells, 980 genes, 17 target cell types plus tumor, 5 patients. Each patient is
one biological replicate. Tumor cells are the query. The question is
which genes in the other cell types change expression with distance to
the nearest tumor cell, reproducibly across the 5 patients.

This vignette follows **paper Figure 3** (A to F), using the corrected
CosMx results bundled with the package. Panel A shows the distance diagnostic;
panels B to F show gradient counts, gene scores, per-patient fits, pathways,
and ligand-receptor candidates. The raw-data preparation and full analysis
are shown but not run when building the vignette.

The required steps are `run_ripple()` and the specificity check.
Everything else is optional. Permutation testing, confounder control,
fGSEA, and ligand-receptor nomination are all downstream stages you
opt into when you think the question calls for them. Each optional step below
includes a brief explanation of why or when to use it

> **Note:** for a 1-minute end-to-end on the synthetic mock dataset
> (no download needed), see
> `vignette("getting_started", package = "ripple")` first. You can use this to just have a look at the functions.

## Dataset preparation (skipped here)

Building the canonical `SpatialExperiment`:

```{r data-prep, eval = FALSE}
library(zellkonverter)
library(SpatialExperiment)

sce <- readH5AD("cosmx_nsclc.h5ad")

# Merge per-patient tumor subtype labels into a single "tumor" class to prepare for the question.
ct <- as.character(colData(sce)$cell_type)
ct[grepl("^tumor", ct)] <- "tumor"

spe <- SpatialExperiment(
  assays        = list(counts = assay(sce, "counts")),
  colData       = DataFrame(
    cell_type = ct,
    patient   = colData(sce)$patient,
    sample    = colData(sce)$sample
  ),
  spatialCoords = as.matrix(reducedDim(sce, "spatial"))
)
```

## Panel A: choosing `k_neighbors` with `plot_k_diagnostics()`

This analysis uses `k = 1`. This makes the gradient score the log-rate change
per micrometer from the nearest tumor cell, up to the 200 µm cap.
For `k > 1`, it instead refers to the mean distance to the nearest `k`
tumor cells. You should inspect the distance curves and gene stability across `k`
when choosing this setting.

```{r k-diagnostics, eval = FALSE}
plot_k_diagnostics(
  input           = spe,
  query_celltype  = "tumor",
  celltype_column = "cell_type",
  sample_column   = "patient",
  k_range         = 1:10,
  max_distance_um = Inf  # inspect uncapped geometry
)
```

The plot averages per-patient mean distances with equal weight per patient for six target
populations, matching Figure 3A.

```{r fig3-A, fig.width = 8, fig.height = 5}
kd <- fread(system.file("extdata/cosmx_cached/k_diagnostics_cosmx.csv",
                        package = "ripple"))
kdiag_show <- c("fibroblast", "macrophage", "endothelial", "neutrophil",
                "T CD8 memory", "B-cell")
kd_mean <- kd[cell_type %in% kdiag_show,
              .(mean_dist = mean(mean_dist)), by = .(k, cell_type)]
ggplot(kd_mean, aes(k, mean_dist, colour = cell_type)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Number of nearest query cells (k)",
       y = expression("Mean distance to tumor (" * mu * "m)"),
       colour = "Target cell type") +
  theme_ripple(base_size = 12)
```

## Running RIPPLE

```{r run-ripple, eval = FALSE}
results <- run_ripple(
  input                = spe,
  query_celltype       = "tumor",
  celltype_column      = "cell_type",
  sample_column        = "patient",
  output_dir           = "ripple_output",
  analysis_name        = "tumor_ripple",
  k_neighbors          = 1,
  max_distance_um      = 200,
  min_cells_per_sample = 30,
  n_permutations       = 0,
  verbose              = TRUE
)
```

Key parameter choices for the main ripple function that you should choose for your datasets:

- **`k_neighbors = 1`**: distance to the single nearest tumor cell;
  appropriate for paracrine signalling range. Run the diagnostic and pick. Alternatively, run the whole analysis with a few different k values.
- **`max_distance_um = 200`**: farther cells havw distance
  capped at 200 µm
- **`sample_column = "patient"`**: each patient is one biological
  replicate (N = 5). Per-patient coefficients are combined via Fisher's
  combined p-value with sign-consistency gating.
- **`n_permutations = 0`**: permutation testing skipped here. It is an
  optional test of query-location specificity. The R API uses non-target
  pseudo-query candidates, it takes a while to run; the standalone GPU script is faster, requires python.

## Loading cached results

The pre-computed outputs are bundled. In your own analysis, replace
these paths with files under your `output_dir/<analysis_name>/`.

```{r load-results}
results_path <- system.file("extdata/cosmx_cached/all_genes_results.csv",
                            package = "ripple")
results <- fread(results_path)

cat(nrow(results), "gene x cell_type combinations tested\n")
cat(sum(results$fisher_fdr < 0.05, na.rm = TRUE), "significant at Fisher FDR < 0.05\n")
```

## Pre-flight: specificity check (run this early!)

> **Strongly recommended:** run the specificity check **before** any
> downstream interpretation, top-hit listing, or pathway enrichment.
> The **broad-expression class** is the set of genes significant in
> many cell types at once. Common offenders: ribosomal genes,
> housekeeping, ambient RNA from highly expressed query markers. If
> you do not filter them, they dominate every "top hits" table and
> every fGSEA leading edge. The cell-type-specific biology gets
> masked. There is an argument for keeping them through analysis and then
> only removing them from interpretation, but it can make things confusing.
>
> **Contamination is user-defined. RIPPLE only proposes candidates.**
> The pipeline gives you three heuristic flags:
>
> 1. The cross-cell-type `broad` class
>    (`classify_gene_specificity()`).
> 2. The ambient-family blocklist (`find_ambient_family_genes()` for
>    Ig, J-chain, Rp, Mt genes). Add genes you already know are ambient
>    in your data, e.g. ones you saw expressed across nearly all
>    populations during cell-type annotation.
> 3. A user-supplied query-marker list (`query_signature_genes` on
>    `run_ripple()`), for genes your query expresses very highly that may
>    spill into neighbours. On 10x Xenium, probes that looked too high
>    versus a reference scRNA-seq are candidates too, confirm they are
>    spilling by checking that reference first.
>
> None of these is a measurement of contamination! A broadly-
> significant gene can be genuine shared biology. Cytokines, MHC II,
> and stress programs are often multi-cell-type. A query
> marker that spills into one neighbouring cell type can be classed
> as `specific` (not `broad`) but still be contamination. See the
> EPCAM-in-fibroblasts case below. You ultimately decide what counts as
> reasonable using domain knowledge. The flags are starting
> points for inspection.

`classify_gene_specificity()` returns one row per significant gene
with two key fields. `n_celltypes` is the number of cell types it
appears in. `specificity_class` is a three-level label: `specific`,
`moderate`, `broad`. We compute it **once**, here, and reuse it
everywhere downstream.

**A note on the broad-class argument.** There is no universally correct
cutoff. `classify_gene_specificity()` exposes **two** dials, both of
which can be edited based on your dataset:

1. **`broad_threshold`** (how many cell types). Default `4` is
   sensible for the ~10-20 cell-type panel here (and for the original LN analyses), but tune to your data:
   - **Few cell types (≤ 5):** a cutoff of 4 is barely reachable;
     consider 2 or 3.
   - **Many fine subtypes (~30+):** a cutoff of 4 is too lenient and
     will flag legitimate lineage-shared biology; consider 6-10.
   - Sanity check: plot `n_celltypes` as a histogram across all
     significant genes. The cutoff should sit in the right tail,
     separating clear multi-cell-type genes from cell-type-specific
     signal.
2. **`broad_sig_threshold`** (how strictly significant each "hit"
   must be). Default `NULL` means any `*`-significant (FDR < 0.05)
   hit counts toward the broad-class tally. Tighten to `"**"`
   (FDR < 0.01) or `"***"` (FDR < 0.001) to make the filter **looser
   overall**: a gene only counts as a hit in cell types where the
   signal is unambiguous, so fewer genes get flagged. Be careful loosening this threshold.

```{r preflight-broad}
spec <- classify_gene_specificity(
  results,
  fdr_threshold       = 0.05,
  broad_threshold     = 4,     # tune for your panel; see note above
  broad_sig_threshold = NULL   # NULL == any "*" hit counts;
                                # use "**" / "***" to require
                                # stricter sig per hit
)

table(spec$specificity_class)

# Genes flagged as broad, to be deprioritised in top-hit lists
broad_genes <- spec[specificity_class == "broad", unique(gene)]
cat("Broad-class genes:", length(broad_genes), "\n")
head(spec[specificity_class == "broad"][order(-n_celltypes)], 10)
```

```{r preflight-clean}
# Filtered copy for top-hit labels. Volcano backgrounds show all genes;
# the cached pathway analysis separately excludes broad-class genes.
results_clean <- results[!gene %in% broad_genes]
cat(nrow(results),       "rows in `results` (unfiltered)\n")
cat(nrow(results_clean), "rows in `results_clean` (broad class removed)\n")
```

> **Limitations of broad-class filtering (read before reporting
> filtered results):**
>
> - **Loses real biology when broad-expression and induction
>   co-occur.** A gene can be both genuinely induced near the query
>   *and* inflated by ambient RNA from query cells. Always inspect the broad-class list and decide
>   case-by-case.
> - **The threshold is panel-dependent.** Default `broad_threshold = 4`
>   assumes 10 to 20 cell types in the panel. With 3 cell types the
>   cutoff is unreachable. With 30 fine subtypes it is too lenient.
>   Re-tune for your annotation granularity.
> - **Removes genes globally, not per cell type.** A gene flagged
>   dataset-wide is dropped from every column. A gene that is
>   broad-expression across most cell types may still be a real
>   signal in the one cell type that biologically expresses it.
> - **Multi-cell-type biology is not ambient RNA.** Cytokines, MHC
>   class II genes, and stress-response programs are
>   expressed across many cell types. They land in the broad class.
>   The heuristic cannot tell them apart from ambient RNA on its own.
> - **Rank-based tests are panel-size sensitive.** Dropping genes
>   shifts the ranks of all remaining genes. NES values from
>   filtered and unfiltered fGSEA runs are not directly comparable.
>   Always report whether the filter was applied. Ideally do both. Doing it on all genes and then filtering post-gsea is a valid option!
>
> **Recommended workflow:**
>
> 1. Classify upfront (this section).
> 2. Keep both `results` (full) and `results_clean` (filtered) in
>    scope.
> 3. Plot all genes as background. Label and report top hits from
>    `results_clean`.
> 4. Run fGSEA both with and without `exclude_broad = TRUE` (Panel E
>    below).

# Reproducing paper Figure 3

The following chunks use the inputs for Figure 3 and demonstrate
the plotting steps. Panel A
is shown above. The volcano below also labels selected flagged genes for
discussion.

## Panel B: gradients per target cell type

```{r fig3-B, fig.width = 9, fig.height = 5}
plot_gene_counts_by_celltype(
  results,
  query_label     = "Tumor",
  broad_threshold = 4
)
```

Each bar is split two ways. By direction: induced near tumor vs
repressed near tumor. By specificity class: solid for cell-type
specific, faded for the "broad" fraction (genes significant in
>= `broad_threshold` cell types). This is the key QC view. A tall bar
only matters biologically if its solid portion dominates. Fibroblasts
and macrophages are the most remodelled cell types near tumor. Both
show a sizeable specific-to-broad ratio.

## Panel C: fibroblast gradient volcano

This is **paper Figure 3C** (fibroblast gradient volcano, IGFBP5
highlighted). The default `plot_gradient_volcano()` works well as a
quick look:

```{r fig3-C-quick, fig.width = 8, fig.height = 6}
plot_gradient_volcano(
  results[cell_type == "fibroblast"],
  query_label = "Tumor",
  n_label     = 15
)
```

For the published figure we colour by direction (red = induced, blue
= repressed) and label a curated gene set. **All** genes still appear as background dots, so the
broad-class genes remain visible for interpretation. But the labels
go on genes that are NOT in the broad list. We add a few flagged
genes by hand (including MIF, IGKC and PTGDS) so we can discuss them.

```{r fig3-C-paper, fig.width = 8, fig.height = 6}
fib <- results[cell_type == "fibroblast"]    # all genes for the scatter
fib[, neg_log10_fdr := -log10(pmax(fisher_fdr, 1e-300))]
fib[, direction := fifelse(
  fisher_fdr < 0.05 & median_coef < 0, "Induced near tumor",
  fifelse(fisher_fdr < 0.05 & median_coef > 0, "Repressed near tumor",
          "Not significant"))]

# Cell-type-specific top hits, pulled from `results_clean` so that
# ambient RNA does not crowd out the cell-type-specific biology.
top_clean_fib <- results_clean[
  cell_type == "fibroblast" & fisher_fdr < 0.05
][order(fisher_fdr)][1:12]$gene

# Plus a few flagged genes we want to discuss explicitly (see "Key
# fibroblast hits" below). MIF, IGKC and RPL22 are broad-class
# (significant in >= 4 cell types); JCHAIN is NOT broad (it is
# significant only in fibroblasts here) but could be an ambient-family gene
# caught by find_ambient_family_genes() instead, or by misslabelling of plasma cells as fibroblasts, or doublets.
discuss_flagged <- c("MIF", "IGKC", "JCHAIN", "RPL22", "PTGDS")

label_fib <- unique(c(top_clean_fib, discuss_flagged))
fib[, to_label := gene %in% label_fib & fisher_fdr < 0.05]

ggplot(fib, aes(x = median_coef, y = neg_log10_fdr)) +
  geom_point(aes(colour = direction), size = 0.9, alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey70") +
  geom_text_repel(
    data = fib[to_label == TRUE], aes(label = gene),
    size = 3.5, max.overlaps = Inf, box.padding = 0.3, seed = 1
  ) +
  scale_colour_manual(values = c(
    "Induced near tumor"   = "#B2182B",
    "Repressed near tumor" = "#2166AC",
    "Not significant"      = "grey75"
  ), name = NULL) +
  labs(
    x = expression("Gradient score (per" ~ mu * "m)"),
    y = expression(-log[10] ~ "Fisher FDR"),
    title = "Fibroblasts: tumor-proximity gene gradients"
  ) +
  theme_ripple(base_size = 12) +
  theme(legend.position = "bottom")
```

**Key fibroblast hits (from `results_clean`):**

- **IGFBP5** (induced): IGF-binding protein involved in stromal
  remodelling; consistent with cancer-associated fibroblast (CAF)
  activation.
- **CXCL12** (repressed): its repression near tumor could reflect phenotypic switching.

**Flagged hits retained for discussion:**

- **MIF** (induced) is significant in seven target cell types and is
  flagged as broad. Its consistency alone cannot distinguish shared
  biology from ambient RNA.
- **PTGDS** (repressed) is significant in four target cell types. It
  remains labelled despite its broad classification.
- **IGKC** and **RPL22** are significant in four and seven target cell
  types, respectively. Their broad and ambient-family flags warrant
  inspection before interpretation.
- **JCHAIN** (repressed) is significant only in fibroblasts, so the
  broad filter misses it. Its plasma-cell association is instead
  flagged by `find_ambient_family_genes()`.

## Panel D: per-patient consistency for IGFBP5

The forest plot shows per-patient gradient scores ± 95% CI for IGFBP5
in fibroblasts. All five slopes are negative under the fitted plateau
model (median `r signif(results[cell_type == "fibroblast" & gene == "IGFBP5", median_coef], 3)`
per µm; Fisher FDR `r format(results[cell_type == "fibroblast" & gene == "IGFBP5", fisher_fdr], scientific = TRUE, digits = 3)`).
IGFBP5 is significant in three target cell types, classified as `moderate`,
and survives the broad-class filter.

```{r fig3-D, fig.width = 7, fig.height = 3.5}
coef_path <- system.file(
  "extdata/cosmx_cached/per_celltype/fibroblast/coef_per_sample.csv",
  package = "ripple"
)
coef_data <- fread(coef_path)

forest_gene <- function(g) {
  d <- coef_data[gene == g]
  d[, ci_lo := coef - 1.96 * se]
  d[, ci_hi := coef + 1.96 * se]
  d
}

igfbp5 <- forest_gene("IGFBP5")

ggplot(igfbp5, aes(x = coef, y = sample_id, colour = sample_id)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(xmin = ci_lo, xmax = ci_hi),
                orientation = "y", width = 0.25, alpha = 0.7) +
  geom_point(size = 3) +
  scale_colour_brewer(palette = "Set2", guide = "none") +
  labs(
    x = expression("Per-patient gradient score for IGFBP5 (per" ~ mu * "m)"),
    y = NULL,
    title = "IGFBP5 in fibroblasts: per-patient gradient scores"
  ) +
  theme_ripple(base_size = 12)
```

All five patients show a consistent negative coefficient: IGFBP5 RNA
counts increase near tumor in every replicate, matching the
CAF-activation biology from Panel C.

### Reproducibility alone is not a contamination check

A unanimous forest plot does not establish a biological mechanism.
MIF also has consistent negative slopes, but is flagged as broad because
it is significant in seven of the 17 target cell types:

```{r fig3-D-mif, fig.width = 7, fig.height = 3.5}
mif <- forest_gene("MIF")

ggplot(mif, aes(x = coef, y = sample_id, colour = sample_id)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(xmin = ci_lo, xmax = ci_hi),
                orientation = "y", width = 0.25, alpha = 0.7) +
  geom_point(size = 3) +
  scale_colour_brewer(palette = "Set2", guide = "none") +
  labs(
    x = expression("Per-patient gradient score for MIF (per" ~ mu * "m)"),
    y = NULL,
    title = "MIF in fibroblasts: per-patient gradient scores"
  ) +
  theme_ripple(base_size = 12)
```

## Panel E: pathway enrichment (fGSEA dotplot)

fGSEA is an **optional** step (why/when: run it to summarise dozens of
significant genes per cell type into interpretable pathway-level
programs).

The bundled Hallmark analysis excludes genes significant in at least four
target cell types (`exclude_broad = TRUE`, `broad_threshold = 4`). It ranks
the remaining genes by median gradient score, using a seed for reproducibility. No additional
query-signature filter was applied to this pathway analysis.

```{r fig3-E-run, eval = FALSE}
fgsea_dt <- run_ripple_fgsea(
  results,
  gene_sets       = "hallmark",
  organism        = "human",
  exclude_broad   = TRUE,
  broad_threshold = 4L,
  seed            = 42
)
```

```{r fig3-E-load}
fgsea_path <- system.file("extdata/cosmx_cached/fgsea_all_celltypes.csv",
                          package = "ripple")
fgsea_dt <- fread(fgsea_path)
```

Dot size encodes -log10(adjusted p-value), and colour encodes NES:
negative NES indicates induction near tumor; positive NES indicates
repression. The axis order is kept consistent with Figure 3E.

```{r fig3-E, fig.width = 9, fig.height = 7}
fig3_cell_order <- c("B-cell", "endothelial", "epithelial", "fibroblast",
  "macrophage", "mast", "mDC", "monocyte", "neutrophil", "NK",
  "pDC", "plasmablast", "T CD4 memory", "T CD4 naive",
  "T CD8 memory", "T CD8 naive", "Treg")
fig3_pathway_order <- c("Interferon Gamma Response", "Allograft Rejection",
  "Adipogenesis", "Glycolysis", "E2f Targets", "G2m Checkpoint")
keep_order <- function(preferred) {
  function(x) c(intersect(preferred, x),
                 sort(setdiff(x, preferred), method = "radix"))
}
plot_fgsea_dotplot(fgsea_dt, padj_threshold = 0.1, top_n = 12) +
  scale_x_discrete(limits = keep_order(fig3_cell_order)) +
  scale_y_discrete(limits = keep_order(fig3_pathway_order))
```


### Checking the effect of filtering

Broad-class filtering can remove genuine shared biology as well as potential
contamination. Compare the leading-edge genes with an unfiltered analysis,
and inspect known query markers. NES values are not directly comparable
between runs that rank different gene sets.

```{r fgsea-filter-comparison, eval = FALSE}
fgsea_unfiltered <- run_ripple_fgsea(
  results, gene_sets = "hallmark", organism = "human",
  exclude_broad = FALSE, seed = 42
)
plot_fgsea_dotplot(fgsea_unfiltered, padj_threshold = 0.1, top_n = 12)

# Optional additional filter, not applied to the Panel E results.
tumor_markers <- c("EPCAM", "KRT8", "KRT18", "KRT19", "KRT7", "KRT17")
fgsea_query_filtered <- run_ripple_fgsea(
  results, gene_sets = "hallmark", organism = "human",
  exclude_broad = TRUE, broad_threshold = 4L,
  query_signature_genes = tumor_markers, seed = 42
)
```

## Panel F: top tumor → fibroblast ligand-receptor pairs

Ligand-receptor nomination is an **optional** step (why/when: run it to
turn a list of gradient genes into testable signalling hypotheses by
matching them to known L-R pairs).

Stage 6 (`run_ripple_lr()`) intersects gradient genes with curated
ligand-receptor databases and (optionally) NicheNet, producing a
combined prioritization score per L-R pair.

By default, `run_ripple_lr()` nominates receptors induced near the query
(negative gradient score). This selects pairs whose receptor expression
increases toward tumor, but it does not establish that the ligand drives that
change. Other receptor directions can be explored explicitly.

```{r fig3-F-run, eval = FALSE}
tumor_markers <- c("EPCAM", "KRT8", "KRT18", "KRT19", "KRT7", "KRT17")
lr_out <- run_ripple_lr(
  results_dir           = "ripple_output/tumor_ripple",
  input                 = spe,
  query_celltype        = "tumor",
  celltype_column       = "cell_type",
  sample_column         = "patient",
  organism              = "human",
  receptor_direction    = "induced",  # default: only receptors UP near tumor
  exclude_contamination = TRUE,       # default: drop cross-cell-type broad receptors
  broad_threshold       = 4L,         # same broad cutoff as the pre-flight
  query_signature_genes = tumor_markers  # drop tumor-marker receptors (e.g. EPCAM)
)
```

The bundled results retain induced receptors after excluding broad-class
receptors and the tumor-marker list above. These filters reduce possible
spillover but do not prove that the remaining pairs are biological.
For example, EPCAM is significant only in fibroblasts here, so it requires
the query-marker filter rather than the broad-class filter.


```{r fig3-F-load}
lr_path <- system.file(
  "extdata/cosmx_cached/lr/fibroblast/combined_prioritization.csv",
  package = "ripple"
)
lr_fib <- fread(lr_path)

cat(nrow(lr_fib), "tumor -> fibroblast L-R pairs (all induced receptors)\n")
head(lr_fib[order(-combined_score),
            .(ligand, receptor, combined_score, receptor_gradient_coef,
              receptor_fdr, ligand_pct_query)], 5)
```

```{r fig3-F, fig.width = 9, fig.height = 6}
# Exclude EZR from receptor candidates, matching the figure selection.
lr_top <- lr_fib[receptor != "EZR"][order(-combined_score)][1:15]
lr_top[, pair := factor(paste0(ligand, " -> ", receptor),
                        levels = rev(paste0(ligand, " -> ", receptor)))]

ggplot(lr_top, aes(x = combined_score, y = pair)) +
  geom_point(aes(size = ligand_pct_query), colour = "#B2182B") +
  scale_size_continuous(range = c(2, 7),
                        name = "Tumor cells\nexpressing ligand (%)") +
  labs(
    x = "Combined prioritization score",
    y = NULL,
    title = "Top tumor -> fibroblast L-R pairs"
  ) +
  theme_ripple(base_size = 12) +
  theme(legend.position = "right",
        legend.box = "vertical",
        legend.key.size = unit(0.4, "cm"))
```

The leading candidates include **COL9A2 → ITGB1**, **COL1A1 → DDR1**,
and **FN1 → ITGB8**, suggesting extracellular-matrix interactions for
follow-up. CD44 → EPCAM is absent after query-marker filtering.
These scores prioritize hypotheses rather than demonstrating signalling.
# Beyond Figure 3: macrophage deep-dive, specificity, and QC

## Macrophage gradient volcano

Macrophages show a polarised response near tumor
(`r sum(results[cell_type == "macrophage", fisher_fdr < 0.05], na.rm = TRUE)`
significant genes). Induced programs: motility, interferon, plus some tumor-RNA
spillover. Repressed programs: antigen-presentation and homeostatic
/ complement. The same volcano view as Panel C, applied to
macrophages:

```{r mac-volcano, fig.width = 8, fig.height = 6}
mac <- results[cell_type == "macrophage"]   # all genes for the scatter
mac[, neg_log10_fdr := -log10(pmax(fisher_fdr, 1e-300))]
mac[, direction := fifelse(
  fisher_fdr < 0.05 & median_coef < 0, "Induced near tumor",
  fifelse(fisher_fdr < 0.05 & median_coef > 0, "Repressed near tumor",
          "Not significant"))]

# Same pattern as the fibroblast panel: top hits from `results_clean`
# plus a curated set of macrophage markers (some, e.g. MHC II, are
# broad-class but kept because they are the biological point here).
# Only fisher_fdr < 0.05 genes are labelled (the `to_label` gate below).
top_clean_mac <- results_clean[
  cell_type == "macrophage" & fisher_fdr < 0.05
][order(fisher_fdr)][1:10]$gene

curated_mac <- c("HLA-DRA", "HLA-DPA1", "HLA-DRB1", "CD74", "HLA-E",
                 "C1QA", "SELENOP", "CD68", "MS4A4A",
                 "S100A6", "MIF", "CXCL10", "TUBB")

label_mac <- unique(c(top_clean_mac, curated_mac))
mac[, to_label := gene %in% label_mac & fisher_fdr < 0.05]

ggplot(mac, aes(x = median_coef, y = neg_log10_fdr)) +
  geom_point(aes(colour = direction), size = 0.9, alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey70") +
  geom_text_repel(
    data = mac[to_label == TRUE], aes(label = gene),
    size = 3.5, max.overlaps = 25, box.padding = 0.3, seed = 1
  ) +
  scale_colour_manual(values = c(
    "Induced near tumor"   = "#B2182B",
    "Repressed near tumor" = "#2166AC",
    "Not significant"      = "grey75"
  ), name = NULL) +
  labs(
    x = expression("Gradient score (per" ~ mu * "m)"),
    y = expression(-log[10] ~ "Fisher FDR"),
    title = "Macrophages: tumor-proximity gene gradients"
  ) +
  theme_ripple(base_size = 12) +
  theme(legend.position = "bottom")
```

**Macrophage induced near tumor (negative gradient score):**

- **S100A6, TUBB** (cytoskeletal / calcium-binding): motility and
  morphological adaptation in the tumor niche.
- **CXCL10**: an interferon-induced chemokine, consistent with an
  inflamed tumor-proximal niche.
- **MIF** (also broad-class): induced in several target types, with
  the same interpretation caveat as in fibroblasts.
- **KRT19** (tumor/epithelial marker): significant in eight target cell
  types and flagged as `broad`. Inspect possible query-marker spillover.

**Macrophage repressed near tumor (positive gradient score):**

- **HLA class II** (HLA-DRA, HLA-DPA1, HLA-DRB1) and **CD74**:
  antigen-presentation machinery is *down* near tumor, consistent with
  reduced MHC II on tumor-proximal macrophages.
- **SELENOP, C1QA, CD68, MS4A4A**: complement and homeostatic /
  tissue-resident macrophage programs, also lower near tumor.

Both the antigen-presentation axis and the homeostatic / complement
axis are repressed near tumor. 

# Specificity breakdown plot and QC dashboard

The specificity check was already run upfront (see
[Pre-flight: specificity check]). The `spec` object and `broad_genes`
vector are still in scope. `plot_specificity_breakdown()` summarises
the specificity classes as a bar plot:

```{r breakdown, fig.width = 7, fig.height = 4}
plot_specificity_breakdown(results, broad_threshold = 4)
```

**HLA-E**, **KRT19**, **MIF**, and **RPL22** are significant in nine,
eight, seven, and seven target cell types, respectively. These broad
patterns warrant inspection but can reflect shared biology as well as
ambient RNA.

## QC dashboard

The compact vignette cache does not include per-cell QC files. Run this
helper on a complete `run_ripple()` output directory to inspect distances,
sample sizes, model results, and query-marker flags.

```{r qc, eval = FALSE}
qc <- ripple_plot_qc(
  results_dir           = "ripple_output/tumor_ripple",
  query_signature_genes = c("KRT19", "EPCAM", "MKI67", "KRT8"),
  query_label           = "Tumor"
)
```

The bleed-through panel (panel 7) takes a query-marker vector. Any
marker that shows up significant in many target cell types is
highlighted in red for inspection of possible ambient RNA.

# Optional stages: confounder control, permutation, diagnostics

The gradient calls above are the core RIPPLE result. The stages in this
section are optional. They are not part of Figure 3; they are additional
controls and diagnostics you opt into when the question calls for them.
None run here (they need the in-memory `spe`, which is not bundled), but
each shows the exact call and how to read the output.

## Stage 4: confounder control

this analysis is optional.
**Why/when:** use it when a second cell type shares the query's niche
and you want to hypothesise whether a gradient is specific to the query or just
reflects the shared neighbourhood. Fibroblasts, macrophages, and
endothelial cells all crowd the tumor-proximal niche here, so a
fibroblast gradient attributed to "distance to tumor" could partly be
"distance to the tumor-adjacent myeloid compartment". `run_ripple_confounder()`
refits a bivariate GLM that adds distance to a control cell type:

```{r stage4-run, eval = FALSE}
stage4 <- run_ripple_confounder(
  input            = spe,
  results_dir      = "ripple_output/tumor_ripple",
  query_celltype   = "tumor",
  celltype_column  = "cell_type",
  sample_column    = "patient",
  control_celltype = "macrophage",  # shares the tumor-proximal niche
  query_label      = "Tumor"
)

table(stage4$classification)
```

Each Stage 1 gene is reclassified from how its query coefficient behaves
once distance-to-control is in the model. The `classification` column
takes the following values (the query-specific label is prefixed with
`query_label`, e.g. `Tumor_specific`):

| Classification | Meaning |
|---|---|
| `Tumor_specific` | Fisher FDR < threshold, same sign as Stage 1, without >10% coefficient growth. |
| `enhanced` | Significant with the same sign and >10% growth in absolute coefficient. |
| `niche_driven` | FDR ≥ threshold and >50% attenuation in absolute coefficient. |
| `underpowered` | FDR ≥ threshold and ≥50% of the absolute coefficient retained. |
| `reversed` | Significant with the opposite sign to Stage 1. |
| `no_stage2_result` | No combined result, including fewer than two valid fits. |
| `unclassified` | Available values do not meet another classification rule. |

Indistinguishable query and control predictors yield `fit_status = "rank_deficient"`
and are excluded from aggregation. `stage2_n_rank_deficient` counts these fits.

Because the bivariate GLM fits only one extra linear distance term, **these
labels are hypothesis-generating**, not a definitive causal decomposition.
Three plotting helpers visualise the output — `plot_confounder_bar()`
(class counts per cell type), `plot_confounder_scatter()` (Stage 1 vs
Stage 4 coefficient), and `plot_confounder_ratio()`:

```{r stage4-plots, eval = FALSE}
plot_confounder_bar(stage4, query_label = "Tumor", control_label = "Macrophage")
plot_confounder_scatter(
  stage4,
  label_genes = c("IGFBP5", "CXCL12", "MMP2"),
  query_label = "Tumor"
)
```

If no suitable control population exists, skip Stage 4. The permutation
test below addresses a different question.

## Stage 2: permutation validation

**Why/when:** compare the observed query locations with random non-target
locations. Set `n_permutations` on the main `run_ripple()` call:

```{r perm-run, eval = FALSE}
results_perm <- run_ripple(
  input           = spe,
  query_celltype  = "tumor",
  celltype_column = "cell_type",
  sample_column   = "patient",
  output_dir      = "ripple_output",
  analysis_name   = "tumor_ripple_perm",
  n_permutations  = 1000,           # adds a perm_pval column
  permutation_pool = "non_target"  # default
)
```

Within each sample, pseudo-query cells are drawn from all non-target cells,
preserving the query count. This tests query-location specificity, not specificity against a particular
alternative cell type.

## Spatial autocorrelation diagnostic

Nearby cells can remain correlated after fitting. Replicate aggregation
does not correct this within-sample dependence. `check_spatial_autocorrelation()`
computes Moran's I on per-sample GLM deviance residuals:

```{r moran, eval = FALSE}
moran <- check_spatial_autocorrelation(
  input           = spe,
  genes           = c("IGFBP5", "CXCL12", "MIF"),
  celltype_column = "cell_type",
  target_celltype = "fibroblast",
  query_celltype  = "tumor",
  sample_column   = "patient",
  k               = 20,   # Moran neighbor graph
  k_neighbors     = 1,    # query-distance setting from the main run
  max_distance_um = 200,
  min_cells_per_sample = 30
)
```
Inspect Moran's I and its p-value per sample; a small I does not establish independence or calibrated Wald
p-values. The diagnostic does not rerun cross-sample gene filtering.

# Interpretation summary

Across 17 target cell types, RIPPLE reports
`r sum(results$fisher_fdr < 0.05, na.rm = TRUE)` significant gene-by-cell-type
combinations involving `r uniqueN(results[fisher_fdr < 0.05, gene])` unique
genes. Reported gradients pass the strict sign gate across their valid
replicates.

- **Fibroblasts:** `r sum(results[cell_type == "fibroblast", fisher_fdr < 0.05], na.rm = TRUE)`
  significant gradients. IGFBP5 is induced near tumor; CXCL12, PTGDS,
  THBS1 and MMP2 are repressed.
- **Macrophages:** `r sum(results[cell_type == "macrophage", fisher_fdr < 0.05], na.rm = TRUE)`
  significant gradients, including induced CXCL10 and repressed
  antigen-presentation genes. SPP1 does not pass the sign gate.
- **Pathways:** broad-filtered Hallmark enrichment identifies proliferation
  and immune-associated gene sets. Inspect their leading-edge genes.
- **Ligand-receptor candidates:** extracellular-matrix pairs lead the
  filtered fibroblast ranking, headed by COL9A2 → ITGB1.

Cross-cell-type flags, ambient-family lists and query markers support
gene-level review. Reproducible RNA gradients nominate associations for
follow-up; they do not establish a signalling mechanism.

## Session info

```{r session-info}
sessionInfo()
```
